---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# jhPlayers

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/jhPlayers)](https://CRAN.R-project.org/package=jhPlayers)
[![R-CMD-check](https://github.com/nim-ach/jhPlayers/workflows/R-CMD-check/badge.svg)](https://github.com/nim-ach/jhPlayers/actions)
<!-- badges: end -->

The goal of jhPlayers is to provide tools for research reproducibility and methods for exploratory data analysis.

## Installation

You can install the development version of jhPlayers from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("nim-ach/jhPlayers")
```

## Example

This is a basic example which shows you how to print mean and standard deviation for specific variables:

```{r example}
library(jhPlayers)

# Compute mean for age, height and weight
colmean_sd(players, c("age", "height", "weight"))
```

We can also compute summary statistics for our data using `summary()` command:

```{r}
summary(players[, 1:4])
```

You can also plot the data, for instance let's plot the relationship between fat-free mass index (ffmi), height and peak power:

```{r, echo = FALSE}
plot(~ ffmi + height + peak_power, players)
```

## Citation

To use `jhPlayers` in publications please use:

```{r}
citation("jhPlayers")
```

